Stack Safety as a Security Property

### 2020-07-09 main.tex
1+2

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
What is stack safety? Numerous mechanisms have been proposed to guard
against attacks that corrupt the call stacks of running programs, and many
others rely on a protected stack as an assumption to provide higher level
guarantees\bcp{this is a good point, but we don't pick it up anywhere in the
  paper, do we?}, a clear and precise definition of stack safety is missing
from the literature.

We propose a formal definition of stack safety phrased in the technical
terminology of language-based security---as a combination of a
confidentiality property (callees should not be able to read their caller's
stack frames) and an integrity property (callees should not corrupt their
caller's stack frames or control state).
  %
We discuss how these properties are enforced or approximated by  
existing stack protection mechanisms and describe some useful
provably stronger and weaker variants.  \bcp{should we mention stktokens here?}
\end{abstract}

@@

\bcp{We don't actually say very clearly anywhere that we are interested in
  low-level enforcement mechanisms for machine-code programs (because this
  is where stack attacks occur, and because at higher levels of abstraction
  the stack is a built-in concept, not something that can be attacked even
  in principle).  This could cause confusion.\apt{not so worried}}

The call stack is a perennial target for attacks
based on memory safety vulnerabilities, with potential consequences
ranging from leakage or corruption of private stack data to
control-flow hijacking. To detect or prevent such attacks, a menagerie of
software and hardware techniques for stack safety have been proposed,
%
including stack canaries~\citep{Cowan+98},
bounds checking~\citep{NagarakatteZMZ09,NagarakatteZMZ10,DeviettiBMZ08},
split stacks~\citep{Kuznetsov+14},
shadow stacks~\citep{Dang+15,Shanbhogue+19},
capabilities~\citep{Woodruff+14,Chisnall+15,Skorstengaard+19,Skorstengaard+19b},
and hardware tagging~\citep{DBLP:conf/sp/RoesslerD18}. \iflater\apt{Mostly from
  nick; there could be more}\fi

But what {\em is} this stack safety of which people speak? Precise formal
definitions are mostly lacking: rather, the notion
is commonly defined in a negative way, by enumerating the bad things that a
given enforcement mechanism
prevents, such as corruption of return addresses, buffer overflows, or
use of uninitialized variables\bcp{Does use of unitialized variables lead to
stack safety attacks?\apt{sure, via buffer overflow}}.
%with the notion usually being
%defined negatively: through examples of stack-based exploits
%\rb{usually though not always; as noted below there are other
%  approaches}\bcp{Yes, we have to be very careful about this}.
%\leo{Examples at a high level}: such as -- does not overwrite return addresses, or ...
Indeed, we are only aware of one quite recent attempt to frame a
positive definition, by
\citet{Skorstengaard+19}, who define stack safety as
the conjunction of two properties: {\em local state encapsulation} and {\em
  well-bracketed control flow}.
They formalize these properties by defining an abstract capability-based machine
that ``satisfies [stack safety] by construction.''
%
While this represents an important step in the right direction, we believe
there is still room to craft a more abstract, more generic characterization
of stack safety that is independent of capabilities or any other specific
enforcement
technique.  In other words, we seek formal criteria that can be applied
to \emph{any} mechanism for protecting stack-based low-level programs to judge
whether or not it actually ``guarantees stack safety.''
%
Our key contribution is a novel, generic presentation of
local state encapsulation and well-bracketed control flow using the
technical framework of language-based security.

@@

\iflater
\bcp{We need to point out somewhere that we are only dealing with
stack safety, though we discuss (somewhere) how our property can be combined
with the heap safety property of [Azevedo et al]}\leo{Should this be
  here or in the assumptions as presented in the running
  example?}\bcp{Probably better there.}
\fi

The intuition behind \citeauthor{Skorstengaard+19}'s {\em well-bracketed control
  flow} is
the need to protect the call graph itself: a callee should pass control
back to the caller only by returning in the standard way (transferring
control to the instruction following the call); it should
not be
allowed to jump back into the caller's code at an arbitrary point, either
directly (via a jump instruction) or indirectly (by corrupting stack
data and returning). Well-bracketed control flow establishes safety
between functions: each instruction is uniquely owned by some
function, and all jumps between functions are either calls that enter
at a marked entry point or returns that re-enter just after their
associated call in the standard nested call structure.  \apt{clarify
  what we are contributing here. We are not trying to phrase this as a security property.}

To phrase these two \apt{which two? Not WBCF!} aspects of stack safety as security properties requires
an explicit attacker model: a clear understanding of who is being protected
and from what. Most existing stack protection mechanisms envisage scenarios
where a callee is influenced by external inputs to corrupt the stack in some
way. To avoid getting into the details of exactly how this happens when
formalizing our properties, we use an even stronger model: when we state the
protections that a function can rely on when it makes a call, we assume that
the callee can execute arbitrary instructions until it returns.

@@

\leo{Took a stab. Needs rewriting}%

After formally defining stack safety, we show that our definition is
enforceable using an existing tag-based enforcement
mechanism~\citep{DBLP:conf/sp/RoesslerD18}. We first consider their
conservative Depth Isolation policy, which enforces the
our basic stack safety property. It also enforces an alternative
lockstep property that requires integrity and confidentiality at every step of
an execution trace, which fails faster and thus is more efficient to test;
we prove (in Coq) that the lockstep property is stronger 
than our basic property,
However, at the same time the
conservative enforcement mechanism is expensive to implement. In the
same work, \citeauthor{DBLP:conf/sp/RoesslerD18} propose a different
mechanism that is more efficient by virtue of being lazy in its
enforcement. At a high level, instead of guaranteeing integrity and
confidentiality when a callee returns, this policy checks it when the caller
actually makes use of (potentially compromised) data. The lazy policy
does not enforce our basic propety, but an enhanced version of it does
enforce a proveably weaker property that still protects
from unsafe callees, but is less intuitive.

In summary, we offer the following contributions:\bcp{I would still like to
  see these points phrased in ``contribution normal form'': say exactly what
  we did for each one (it's a bit vague---''considers,'' etc.), and
  explicitly say why it is (a) useful, (b) challenging, and (c) novel.}
\begin{itemize}
\item
  % \leo{More? Extensional}\bcp{We might need to invent a word for   this---''quasi-extensional''?}
  A novel formal stack safety property capturing the
  intuitive concepts of strict local state encapsulation and well-bracketed
  control flow~(\cref{sec:lse-and-wbcf}) in terms of security concepts.
  % These properties   rule out common examples of stack memory abuse.
  This property is enforced by 
  the Depth Isolation policy of \citet{DBLP:conf/sp/RoesslerD18}, based on hardware tags~(\cref{sec:conservative}).
\item
  A lockstep stack safety property that is also enforced by the Depth Isolation policy,
  and is more efficient to test, which we prove (in Coq) is stronger than the basic property~(\cref{sec:lockstep}).
\item
  A novel lazy stack safety property that is enforced by the more efficient
  Lazy Tagging and Clearing policy
  of \citet{DBLP:conf/sp/RoesslerD18}, which we prove (in Coq) is weaker than the basic property~(\cref{sec:lazy}).
%  This policy does not enforce our strict
%  local state encapsulation property,
%  but a modified version of it using unique per-activation tags enforces
%  an (also provably) weaker lazy variant of the property, which still prohibits
%  the same intuitively stack unsafe behaviors.
%
% \apt{Awkward, but the truth I fear.} \bcp{I agree the phrasing is awkward,
%   but the point we can make is still strong: To make the property easier
%   to enforce, we can slightly weaken its guarantees; the weaker version is
%   a bit more complicated to state but still retains the pragmatic
%%   protections of the original.}\apt{To clarify: Nick's LT\&C policy uses
%   depth-based tags, which do not enforce even our lazy policy. So we need to
%   introduce a version of LT\&C that uses unique activation tags.}

@@

To make our stack safety definitions precise, we make some simple assumptions about
memory layout; our definitions could easily be changed to be parametric over
these assumptions, at the cost of slightly greater complexity.  Memory 
consists of an instruction region and a stack that grows upwards; for simplicity,
we omit a heap. (Integrating a memory safety property for the heap in the style of
\citet{DBLP:conf/post/AmorimHP18} is a natural extension that we discuss
in~\Cref{sec:related}.)  The stack has a conventional structure, with
contiguously allocated per-function frames.
The boundary between the frames of a caller and callee is given by the $\rsp$ register
at the time of the call.  The stack can be used to
pass arguments or return results in the usual way, by placing them
at the top of the caller's frame. Functions can also communicate via registers; indeed, our 
definitions say nothing at all about registers (except $\PCname$ and $\rsp$).

@@

We model ordinary program state and policy enforcement state separately to make it 
easier to use well-established concepts from language-based security.  
In particular, confidentiality has an elegant extensional characterization in terms of
\emph{noninterference}: a program preserves the confidentiality of certain data if
\emph{varying} that data would not change the program's observable behavior~\citep{??}.
In applying this idea to a system that incorporates policy enforcement, only
ordinary program data should be varied, whereas any associated policy enforcement
state should be left alone; our model makes this distinction obvious.
%That is, we compare the actual execution of the program with a \emph{hypothetical}
%execution where data has been varied. 
We also introduce a notion of observational integrity that relies on the idea of
\emph{rolling back} program state; again, our model makes it easy to describe
rollback of ordinary program data while leaving policy enforcement data unchanged.

@@

\paragraph*{Capability machines}
%
%Other policies are implemented in hardware or built on hardware mechanisms.
Capability machines like CHERI \citep{Woodruff+14} extend conventional
architectures to support efficient and fine-grained control over memory
accesses, which can be used to implement security policies, including 
the stack protection policies of
\citet{Skorstengaard+19,Skorstengaard+19b}.  Capabilities tightly package pointers
(ordinary data) with base and bounds information (policy data),
and their use requires rewriting source code (e.g. to use capabilities in place
of ordinary pointers), so separating ordinary machine state from policy
state is delicate. \apt{But we conjecture it can be done...?}\leo{Just say that?}

@@

\bcp{We really want readers to understand this trace.  I think we should
  spend a few more words walking through it in detail.}

@@

To streamline the example, we use a small instruction set reminiscent of
RISC-V and make a few simplifying assumptions about the organization of
memory.
%
First, since our goal is to formalize stack safety, we ignore the heap
and assume that all variables are stack allocated. As noted in the
previous section, the extension to account for memory safety is
discussed in~\Cref{sec:related}.
%
We also assume that the machine is word- (not byte-) addressed, that program
instructions start from location 0, that the machine communicates with the
outside world through a single memory-mapped output port at address 99 (so
a {\tt print} statement will be compiled to a store to this
location), and that the stack begins at location 100 and grows upwards.
% \leo{Explain more?  How to connect to realistic one? Observations? DMA?
%   Interrupt handlers?}\bcp{Yes.}

>>>>

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
What is stack safety? Numerous mechanisms have been proposed to guard
against attacks that corrupt the call stacks of running programs, and many
others rely on a protected stack as an assumption to provide higher level
guarantees\bcp{this is a good point, but we don't pick it up anywhere in the
  paper, do we?}\leo{No, should we delete this? it still makes sense in flow}, a clear and precise definition of stack safety is missing
from the literature.

We propose a formal definition of stack safety phrased in the technical
terminology of language-based security---as a combination of a
confidentiality property (callees should not be able to read their caller's
stack frames) and an integrity property (callees should not corrupt their
caller's stack frames or control state).
  %
We discuss how these properties are enforced or approximated by  
existing stack protection mechanisms and describe some useful
provably stronger and weaker variants.  \bcp{should we mention stktokens here?}
\leo{Not sure. We almost never cite stuff in the abstract, do we?}
\end{abstract}

@@

%\bcp{We don't actually say very clearly anywhere that we are interested in
%  low-level enforcement mechanisms for machine-code programs (because this
%  is where stack attacks occur, and because at higher levels of abstraction
%  the stack is a built-in concept, not something that can be attacked even
%  in principle).  This could cause confusion.\apt{not so worried}}
%\leo{I agree. I've clarified the writing in the top of the running example
%  a bit to make sure though}

The call stack is a perennial target for attacks
based on memory safety vulnerabilities, with potential consequences
ranging from leakage or corruption of private stack data to
control-flow hijacking. To detect or prevent such attacks, a menagerie of
software and hardware techniques for stack safety have been proposed,
%
including stack canaries~\citep{Cowan+98},
bounds checking~\citep{NagarakatteZMZ09,NagarakatteZMZ10,DeviettiBMZ08},
split stacks~\citep{Kuznetsov+14},
shadow stacks~\citep{Dang+15,Shanbhogue+19},
capabilities~\citep{Woodruff+14,Chisnall+15,Skorstengaard+19,Skorstengaard+19b},
and hardware tagging~\citep{DBLP:conf/sp/RoesslerD18}. \iflater\apt{Mostly from
  nick; there could be more}\leo{Do we want to hunt for more refs here?}\fi

But what {\em is} this stack safety of which people speak? Precise formal
definitions are mostly lacking: rather, the notion
is commonly defined in a negative way, by enumerating the bad things that a
given enforcement mechanism
prevents, such as corruption of return addresses, buffer overflows, or
use of uninitialized variables.
%\bcp{Does use of unitialized variables lead to stack safety attacks?\apt{sure, via buffer overflow}}
%with the notion usually being
%defined negatively: through examples of stack-based exploits
%\rb{usually though not always; as noted below there are other
%  approaches}\bcp{Yes, we have to be very careful about this}.
%\leo{Examples at a high level}: such as -- does not overwrite return addresses, or ...
Indeed, we are only aware of one quite recent attempt to frame a
positive definition, by
\citet{Skorstengaard+19}, who define stack safety as
the conjunction of two properties: {\em local state encapsulation} and {\em
  well-bracketed control flow}.
They formalize these properties by defining an abstract capability-based machine
that ``satisfies [stack safety] by construction.''
%
While this represents an important step in the right direction, we believe
there is still room to craft a more abstract, more generic characterization
of stack safety that is independent of capabilities or any other specific
enforcement
technique.  In other words, we seek formal criteria that can be applied
to \emph{any} mechanism for protecting stack-based low-level programs to judge
whether or not it actually ``guarantees stack safety.''
%
Our key contribution is such a novel and generic presentation of
local state encapsulation and well-bracketed control flow using the
technical framework of language-based security.

@@

The intuition behind \citeauthor{Skorstengaard+19}'s {\em
  well-bracketed control flow} is the need to protect the call graph
itself: a callee should pass control back to the caller only by
returning in the standard way (transferring control to the instruction
following the call); it should not be allowed to jump back into the
caller's code at an arbitrary point, either directly (via a jump
instruction) or indirectly (by corrupting stack data and
returning). Well-bracketed control flow establishes safety between
functions: each instruction is uniquely owned by some function, and
all jumps between functions are either calls that enter at a marked
entry point or returns that re-enter just after their associated call
in the standard nested call structure.  \apt{clarify what we are
  contributing here. We are not trying to phrase this as a security
  property.}\leo{Better now? Should this be fixed two paragraphs above too?}

To phrase stack safety, and especially local state encapsulation, as a
security property requires an explicit attacker model: a clear
understanding of who is being protected and from what. Most existing
stack protection mechanisms envisage scenarios where a callee is
influenced by external inputs to corrupt the stack in some way. To
avoid getting into the details of exactly how this happens when
formalizing our properties, we use an even stronger model: when we
state the protections that a function can rely on when it makes a
call, we assume that the callee can execute arbitrary instructions
until it returns.

@@

After formally defining stack safety, we show that our definition is
enforceable using an existing tag-based enforcement
mechanism~\citep{DBLP:conf/sp/RoesslerD18}. We first consider their
conservative Depth Isolation policy, which not only enforces our basic
stack safety property, but also an alternative lockstep property that
checks for integrity and confidentiality violations at every step of
an execution trace. Thus, this property fails faster and can lead to
more efficient testing. We prove (in Coq) that this lockstep property
implies our basic definition of stack safety.

However, at the same time the conservative enforcement mechanism is
expensive to implement. In the same work,
\citeauthor{DBLP:conf/sp/RoesslerD18} propose a different mechanism
that is more efficient by virtue of being lazy in its enforcement. At
a high level, instead of guaranteeing integrity and confidentiality
when a callee returns, this policy checks for violations when the
caller actually makes use of (potentially compromised) data. The lazy
policy does not enforce our basic propety, but an enhanced version of
it does enforce a proveably weaker property that still protects from
unsafe callees, but is less intuitive.

In summary, we offer the following contributions:
%\bcp{I would still like to
%  see these points phrased in ``contribution normal form'': say exactly what
%  we did for each one (it's a bit vague---''considers,'' etc.), and
%  explicitly say why it is (a) useful, (b) challenging, and (c) novel.}
% +1 on Andrew's rewrite
\begin{itemize}
\item
  % \leo{More? Extensional}\bcp{We might need to invent a word for   this---''quasi-extensional''?}
  A novel formal stack safety property capturing the
  intuitive concepts of strict local state encapsulation and well-bracketed
  control flow~(\cref{sec:lse-and-wbcf}) in terms of security concepts.
  % These properties   rule out common examples of stack memory abuse.
  This property is enforced by 
  the Depth Isolation policy of \citet{DBLP:conf/sp/RoesslerD18}, based on hardware tags~(\cref{sec:conservative}).
\item
  A lockstep stack safety property that is also enforced by the Depth Isolation policy,
  and is more efficient to test, which we prove (in Coq) is stronger than the basic property~(\cref{sec:lockstep}).
\item
  A novel lazy stack safety property that is enforced by the more efficient
  Lazy Tagging and Clearing policy
  of \citet{DBLP:conf/sp/RoesslerD18}, which we prove (in Coq) is weaker than the basic property~(\cref{sec:lazy}).
%  This policy does not enforce our strict
%  local state encapsulation property,
%  but a modified version of it using unique per-activation tags enforces
%  an (also provably) weaker lazy variant of the property, which still prohibits
%  the same intuitively stack unsafe behaviors.
%
% \apt{Awkward, but the truth I fear.} \bcp{I agree the phrasing is awkward,
%   but the point we can make is still strong: To make the property easier
%   to enforce, we can slightly weaken its guarantees; the weaker version is
%   a bit more complicated to state but still retains the pragmatic
%%   protections of the original.}\apt{To clarify: Nick's LT\&C policy uses
%   depth-based tags, which do not enforce even our lazy policy. So we need to
%   introduce a version of LT\&C that uses unique activation tags.}

@@

To make our stack safety definitions precise, we make some simple assumptions about
memory layout; our definitions could easily be changed to be parametric over
these assumptions, at the cost of greater complexity.  Memory 
consists of an instruction region and a stack that grows upwards; for simplicity,
we omit a heap. (Integrating a memory safety property for the heap in the style of
\citet{DBLP:conf/post/AmorimHP18} is a natural extension that we discuss
in~\Cref{sec:relwork}.)  The stack has a conventional structure, with
contiguously allocated per-function frames.
The boundary between the frames of a caller and callee is given by the $\rsp$ register
at the time of the call.  The stack can be used to
pass arguments or return results in the usual way, by placing them
at the top of the caller's frame. Functions can also communicate via registers; indeed, our 
definitions say nothing at all about registers (except $\PCname$ and $\rsp$).

@@

We model ordinary program state and policy enforcement state separately to make it 
easier to use well-established concepts from language-based security.  
In particular, confidentiality has an elegant extensional characterization in terms of
\emph{noninterference}: a program preserves the confidentiality of certain data if
\emph{varying} that data would not change the program's observable behavior.\leo{What citation should go here? Not the hyperproperties one?}
In applying this idea to a system that incorporates policy enforcement, only
ordinary program data should be varied, whereas any associated policy enforcement
state should be left alone; our model makes this distinction obvious.
%That is, we compare the actual execution of the program with a \emph{hypothetical}
%execution where data has been varied. 
We also introduce a notion of observational integrity that relies on the idea of
\emph{rolling back} program state; again, our model makes it easy to describe
rollback of ordinary program data while leaving policy enforcement data unchanged.

@@

\paragraph*{Capability machines}
%
%Other policies are implemented in hardware or built on hardware mechanisms.
Capability machines like CHERI \citep{Woodruff+14} extend conventional
architectures to support efficient and fine-grained control over memory
accesses, which can be used to implement security policies, including 
the stack protection policies of
\citet{Skorstengaard+19,Skorstengaard+19b}.  Capabilities tightly package pointers
(ordinary data) with base and bounds information (policy data),
and their use requires rewriting source code (e.g. to use capabilities in place
of ordinary pointers), so separating ordinary machine state from policy
state is delicate, but we conjecture it can be done with some effort.
%\apt{But we conjecture it can be done...?}\leo{Just say that?}

@@

\bcp{We really want readers to understand this trace.  I think we should
  spend a few more words walking through it in detail.}\leo{Not sure
  exactly what you mean here. It would take roughly a page to go through
  each instruction in detail - that's what the last column is for in my mind.
  Or do you have something else in mind?}

@@

To streamline the example, we use a small instruction set reminiscent of
RISC-V and make a few simplifying assumptions about the organization of
memory.
%
First, since our goal is to formalize stack safety, we ignore the heap
and assume that all variables are stack allocated. As noted in the
previous section, the extension to account for memory safety is
discussed in~\Cref{sec:relwork}.
%
We also assume that the machine is word- (not byte-) addressed, that program
instructions start from location 0, that the machine communicates with the
outside world through a single memory-mapped output port at address 99 (so
a {\tt print} statement will be compiled to a store to this
location), and that the stack begins at location 100 and grows upwards.
% \leo{Explain more?  How to connect to realistic one? Observations? DMA?
%   Interrupt handlers?}\bcp{Yes.}
